When does vapor pressure deficit drive or reduce evapotranspiration?

### 2017-10-12 doc/paper/vpd_et_paper.tex
finished with superficial comments up to page 7, need to start working on meatier conecptual comments, read monteith's oroginal 1965 paper.

Changes to vapor pressure deficit (VPD) alter the atmospheric demand for water from the land surface. However, plant stomata have evolved to optimally regulate the exchange of water and carbon between vegetation and the atmosphere \citep{Franks_2017}. Therefore, an increase (decrease) in VPD may not correspond to an increase (decrease) in evapotranspiration (ET) because stomatal closure (opening) can cancel the effects of shifts to atmospheric demand.

Quantifying the plant response to a perturbation to atmospheric VPD increases our understanding of feedbacks between the land surface and the atmosphere. If plant response reduces ET in response to an increase in VPD, the land surface will contribute a positive feedback in response to atmospheric drying. Conversely, if plant response increases ET in response to increase in VPD, then the land surface will contribute a negative feedback to atmospheric drying. The sign of these feedbacks drives the evolution of the atmosphere and land-surface at many timescales, from diurnal to interdecadal.

Here we use a Penman-Monteith framework to quantify plant response to perturbations to atmospheric demand for water. Section 2 derives the framework, Section 3 describes the data used, Section 4 presents results, and Section 5 discusses conclusions. The goal of this paper is to use reasonable approximations as a tool to increase intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity climate models.

@@

This can be adapted to an ecosystem-scale stomatal resistance by multiplying by leaf area index (LAI) and converting units to m s$^{-1}$:
\begin{linenomath*}
  \begin{equation}
  g_s = \text{LAI } \frac{R* T}{P} \left( g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{D_{s}}}\right) \frac{A}{c_s}\right)
  \end{equation}
\end{linenomath*}

While Equation 3 can be used in PM, it will make analytical work with the function intractable because $A$ is a relatively strong function of ET. To remove dependence of ET on $A$ we can use the semi-empirical results of \citet{Zhou_2015}. \citet{Zhou_2015} showed that:

@@

\begin{linenomath*}
  \begin{equation}
  g_s = \text{LAI } \frac{R* T}{P} 1.6 \left(1 + \frac{g_1}{\sqrt{D_{s}}}\right) \frac{uWUE \; ET}{c_s \; \sqrt{D}}
  \end{equation}
\end{linenomath*}

Plugging Equation 5 into Equation 1 and rearranging gives:

\begin{linenomath*}
  \begin{equation}
  ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p D_{s}}{R_{air}} - \frac{\gamma c_s \sqrt{D} }{\text{LAI } R* \; 1.6 \text{ uWUE } (1 + \frac{g_1}{\sqrt{D}})} \right) }{ \Delta + \gamma}
  \end{equation}
\end{linenomath*}

We can then take the derivative with respect to $D$ to determine ecosystem response to atmospheric demand perturbations:

\begin{linenomath*}
  \begin{equation}
    \frac{\partial \;  ET}{\partial \; D} = \frac{g_a \; P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} - \frac{\gamma c_s }{\text{LAI }1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{D}}{2 (g_1 + \sqrt{D})^2}\right) \right)
    \label{d_et}
  \end{equation}
\end{linenomath*}
Note that given yearly uWUE from \citet{Zhou_2015}, $g_1$ from \citet{Lin_2015} \citep[as presented in ][]{Franks_2017}, and observations of R, T, P, D$_s$, and wind speed (WS), the only unknown is LAI. With flux tower observations of ET, LAI will then be uniquely determined for each observation through Equation 6:

\begin{linenomath*}
  \begin{equation}
    LAI  = - \frac{g_a \gamma c_s \sqrt{D_s} P }{ \left(\text{ ET } ( \Delta + \gamma) - \Delta R - g_a \rho_a c_p D_{s}\right) 1.6 \; R\; T\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{D_s}})}
    \label{lai}
  \end{equation}
\end{linenomath*}

This ``pseudo-LAI'' is some part ``true'' LAI (a measure of leaf area), and some part model and observational error, including error involving our assumption of constant uWUE. By calculating a unique LAI for each observation we will propagate any model and observational uncertainty forward into our expression for $\frac{\partial \; ET}{\partial D}$.

\begin{table}
\caption{Definition of symbols and variables}
\centering
\begin{tabular}{l c c}
\hline
 Variable & Description & Units  \\
\hline
$e_s$  & saturation vapor pressure & Pa  \\ 
$T$  & temperature  & K \\
$\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ \\
$R$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   \\
  $g_a$  & atmospheric conductance & m s$^{-1}$  \\
  $\rho_a$  & air density & kg m$^{-3}$  \\
  $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ \\
  $D$  & VPD & Pa  \\
  $\gamma$  & psychrometric constant & Pa K$^{-1}$   \\
  $g_s$  & stomatal conductance & m s$^{-1}$  \\
  $g_{l-s}$  & leaf-scale stomatal conductance & mol m$^{-2}$ s$^{-1}$  \\
  $R*$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ \\
  $LAI$ & leaf area index & -\\
  $c_s$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air\\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

@@

By construction, the variability in the LAI term (Equation \ref{lai}) contains all model and observational uncertainties. LAI also has physical meaning corresponding to ``true'' leaf area, and we expect that it would be approximately $O(1)$, especially because uWUE is fit in \citet{Zhou_2015} at the ecosystem-level. We can have some confidence in our framework, including the assumption of constant uWUE, if calculated LAIs are generally $O(1)$. Figure \ref{lai_fig} presents the histogram of calculated LAIs with outliers (lowest and highest 5\% percent) and nonphysical values (LAI $<$ 0.) removed. All remaining LAI values are $O(1)$ which provides confidence in model framework.

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig02.png}
\caption{Histogram of LAI values calculated for each site and time according to Equation \ref{lai}. The lowest and highest 5\% are removed as outliers, as well as any values below 0. The curve is normalized such that its area is 1. }
\label{lai_fig}
\end{figure}

An additional concern is that the LAI term may in fact be some function of $D$, in which case the dependence would need to be accounted for when taking the derivative. Figure \ref{lai_vpd_fig}\explain{Figs 2 and 3 can probably be combined - the histogram of LAI is shown in Fig 3}. plots the joint distribution of LAI and VPD, and shows that LAI is very weakly a function of VPD. Given this weak dependence, we argue that Equation \ref{d_et} is a valid approximation for ET response to $D$.

% below show histogram of site specific spearmannr? think would be useful 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $D$ and LAI. LAI has only a weak dependence on $D$. ***This plot could proabably benefit from a box plot of site specific correlations, because some sites do have stronger depednence than others. Note also Figs 3 and 2 can probably be combined because this figure shoes LAI's histogram.***}
\label{lai_vpd_fig}
\end{figure}

Before diving into calculated values of $\frac{\partial \; ET}{\partial \; D}$, it is useful to consider the functional form of Equation \ref{d_et}. There are three terms: a scaling term for the full expression we will call Term 1 ($\frac{g_a \; P}{T(\Delta + \gamma)}$), a relatively constant offset we will call Term 2 ($\frac{c_p}{R_{air}}$), and a variable term we will call Term 3 ($\frac{\gamma c_s }{\text{LAI }1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{D}}{2 (g_1 + \sqrt{D})^2}\right)$). All variables are positive, so the relative magnitude between Term 2 and Term 3 will determine the sign of the derivative, while Term 1 will scale the expression larger or smaller.

@@

Term 2 minus Term 3 determines the sign and magnitude that the scaling Term 1 is multiplied by. If we assume that $c_s$ variability is relatively less than LAI and $D$ variability, then variability within PFT will be solely determined by LAI and $D$. Figure \ref{term3} shows how (Term 2 - Term 3) varies with $D$ and LAI, as a function of PFT. In Figure \ref{term3}a lower uWUE and LAI shift the distribution of (Term 2 - Term 3) towards negative values. Additionally, the smaller $g1$, the greater the relative $D$ dependence of of (Term 2 - Term 3). This is observed most strongly for the ENF PFT, which has the smallest $g1$ (74.31).

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for Term 2 - Term 3. Top: Term 2 - Term 3 as a function of VPD, with LAI held fixed at PFT averages. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and LAI. Lines and stars denote the distribution of VPD and LAI next to each axis, following the same percentiles as above.}
\label{term3}
\end{figure}

Figure \ref{term3}b shows the location of the minima of ET, as a function of LAI and $D$. For any LAI or VPD less (more) than these curves, Term 2 - Term 3 will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $LAI$. However, we can see some general trends. For CSH, $\frac{\partial \; ET}{\partial \; D}$ should be negative for the vast majority of observed LAI and VPD. The split appears to be more even among ENF, GRA, and DBF, and we might expect a greater frequency of positive $\frac{\partial \; ET}{\partial \; D}$ for CRO.

Table 3 confirms these expectations for PFT behavior of $\frac{\partial \; ET}{\partial \; D}$. For all PFTs except for CRO, average $\frac{\partial \; ET}{\partial \; D}$ is less than zero. However, $\frac{\partial \; ET}{\partial \; D}$ evaluated at the average of all variables (e.g. LAI, $T$, $c_s$, $D$) is only negative for CSH and GRA. And, DBF in addition to CRO experiences $\frac{\partial \; ET}{\partial \; D}$ < 0 less than half the time. These observations highlight the effect of the nonlinear function in Figure \ref{term3}: $\frac{\partial \; ET}{\partial \; D}$ has a much steeper slope when the function is negative, and is thus more likely to be large magnitude.

The units of $\frac{\partial \; ET}{\partial \; D}$ make it difficult to interpret if $D$ is even a meaningful contributor to ET's variability. To understand $D$'s contribution better, we use a linear approximation and present $\frac{\partial \; ET}{\partial \; D}$ multiplied by $D$'s standard deviation. The $D$'s contribution to ET's variability ranges between 16 - 20 W m$^{-2}$ for all PFTs except for CSH, which is about 51 W m$^{-2}$. Another meaningful comparison is to $\frac{\partial \; ET}{\partial \; R} * std(R)$, as net radiation generally the driver of ET (cite joe berry here). For all PFTs except for CSH $D$ contributes between 14.5 - 20.5 \% of $R$'s contribution to variability. For CSH the portion is much larger, about 44 \%. However it is important to note that a linear approximation about a mean base state is probably not a very good approximation across the range of variability, so these values are just estimates of $D$'s contribution to ET's variability. Regardless, $D$'s variability is certiantly a non-negligable contributor to $ET$'s variability.

So far, idealized plots and statistics have illuminated the form of $\frac{\partial \; ET}{\partial \; D}$ and how it varies with PFT. Large mean LAI and uWUE shifts CRO and DBF towards positive $\frac{\partial \; ET}{\partial \; D}$. However, the strongly nonlinear function of $\frac{\partial \; ET}{\partial \; D}$ at $\frac{\partial \; ET}{\partial \; D} < 0$ pushes $\overline{\frac{\partial \; ET}{\partial \; D}}$ negative for DBF (it does not do this for CRO because of CRO's high $g1$). ENF's low $g1$ value increases the dependence of $\frac{\partial \; ET}{\partial \; D}$ on $D$, and makes the function more strongly nonlinear. This has the side effect of pushing $\overline{\frac{\partial \; ET}{\partial \; D}}$ negative further than other PFTs for a given fraction $\frac{\partial \; ET}{\partial \; D} < 0$ and magnitude $\frac{\partial \; ET}{\partial \; D}(\overline{T,\ldots,D})$. GRA shows the opposite behavior; a relatively high $g1$  makes the function more linear, decreasing the magnitude of $-\overline{\frac{\partial \; ET}{\partial \; D}}$ for a given  [large] fraction $\frac{\partial \; ET}{\partial \; D} < 0$ and negative $\frac{\partial \; ET}{\partial \; D}(\overline{T,\ldots,D})$ (although $g_a$ and Term 1 also probably have a role in this). Finally, low $uWUE$ of CSH pushes to toward by far the lowest values $\frac{\partial \; ET}{\partial \; D}$ (Figure \ref{term3}). Variability in $D$ accounts for the largest about of $ET$ variability for CSH. For the other PFTs, $D$ contributes less to $ET$ variability, but still represents about 15-20 \% of $R$'s contributions to ET variability.

@@

The full observations generally confirm expectations from Section \ref{results}. CRO has the most positive values of $\frac{\partial \; ET}{\partial \; D}$, $\frac{\partial \; ET}{\partial \; D}$ is almost always negative for CSH, and response depends more with the environmental conditions for the other PFTs (especially ENF). Through the columns of Figure \ref{real} we can see the impact of LAI and $g_a$ on the variability of $\frac{\partial \; ET}{\partial \; D}$. $g_a$'s scaling (included in columns 1 and 3) alters the magnitude considerably. $LAI$ variability (included in columns 1 and 2) adds a lot of additional noise to the signal of $\frac{\partial \; ET}{\partial \; D}$, which is slightly undesirable given LAI's role in representing model and observational uncertainty. However, the general story with the noise appears to match the cleaner signal when $LAI$ is help constant and $D_{ETmin}$ is clearly visible \explain{I really need to make these plots better - way too much overlapping of points that hurts the story}. One exception is possibly with GRA, for which uncertainty represented in LAI is high and causes the full complexity plots (Columns 1 and 2) to not match well with LAI held fixed (Columns 3 and 4).

For ENF and GRA $D_{ETmin}$ does not appear to be only a function of LAI (most observable in Column 4). It turns out that the site to site variability in $\gamma$ causes $D_{ETmin}$ to vary, which is not discussed in the previous section. The impact is observable in both ENF and GRA, but especially for ENF which has a larger $\frac{\partial^2 \; ET}{\partial^2 \; D}$ than the other PFTs.

>>>>

Changes to vapor pressure deficit (VPD) alter the atmospheric demand for water from the land surface. Traditionally, atmospheric scientists and hydrometeorologists generally think that an increase in atmospheric demand induces an increases in evapotranspiration (ET) (citations?). This possible misconception developed in part due to the proliferation of studies examing potential ET rather than estimates of ET itself (citations?).  In contrast, plant physiologists know that stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase (decrease) in VPD may not correspond to an increase (decrease) in ET because stomatal closure (opening) can cancel the effects of shifts to atmospheric demand.

Quantifying the plant response to a perturbation to atmospheric VPD increases our understanding of land surface response to shifts in atmospheric conditions. If plant response reduces ET in response to atmospheric drying then soil moisture will be conserved. An increase in ET in reponse to atmospheric drying will reduce soil moisture, but contribute increased moisenting to the atmosperhere. Clearly, the sign and magnitude of land-surface responsedrives the co-evolution of the atmosphere and land-surface at many timescales, from diurna to interdecadal.

In order to quantify plant response to perturbations to atmospheric demand for water, we apply a Penman-Monteith framework to eddy-covariance observation spanning various biomes and climates. Section 2 describes the data used, Section 3 derives the framework, Section 4 presents results, and Section 5 discusses conclusions. The goal of this paper is to use reasonable approximations as a tool to increase intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity climate models.

@@

This can be adapted to an ecosystem-scale stomatal resistance by multiplying by leaf area index (LAI) and converting units to m s$^{-1}$:
\begin{linenomath*}
  \label{medlyn}
  \begin{equation}
  g_s = \text{LAI } \frac{R \,T}{P} \left( g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{D_{s}}}\right) \frac{A}{c_s}\right)
  \end{equation}
\end{linenomath*}

While Equation 3 can be used in PM, it will make analytical work with the function intractable because $A$ is a function of ET itself. To remove dependence of ET on $A$ we can use the semi-empirical results of \citet{Zhou_2015}. \citet{Zhou_2015} showed that:

@@

\begin{linenomath*}
  \begin{equation}
  g_s = \frac{\text{LAI }}{\text{LAI$_{ref}$}} \frac{R \, T}{P} 1.6 \left(1 + \frac{g_1}{\sqrt{D_{s}}}\right) \frac{uWUE \; ET}{c_s \; \sqrt{D}}
  \end{equation}
\end{linenomath*}

Note that $uWUE$ is fit on the ecosystem scale in \citet{Zhou_2015} so it already has an implicit LAI$_{ref}$ reflective of the average $LAI$ in \citet{Zhou_2015}'s fits. So the LAI term in Equation \ref{medlyn} becomes a term we call $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$, which is a measure of how close the observed LAI is to the LAI$_{ref}$ in \citet{Zhou_2015}.

Next, plugging Equation 5 into Equation 1 and rearranging gives:

\begin{linenomath*}
  \begin{equation}
  ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p D_{s}}{R_{air}} - \frac{\text{LAI$_{ref}$}}{\text{LAI}}  \frac{\gamma c_s \sqrt{D} }{ R* \; 1.6 \text{ uWUE } (1 + \frac{g_1}{\sqrt{D}})} \right) }{ \Delta + \gamma}
  \end{equation}
\end{linenomath*}

We can then take the derivative with respect to $D$ to analytically determine ecosystem response to atmospheric demand perturbations:

\begin{linenomath*}
  \begin{equation}
    \frac{\partial \;  ET}{\partial \; D} = \frac{g_a \; P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} - \frac{\text{LAI$_{ref}$}}{\text{LAI}} \frac{\gamma c_s }{1.6 \; R*\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{D}}{2 (g_1 + \sqrt{D})^2}\right) \right)
    \label{d_et}
  \end{equation}
\end{linenomath*}
Note that given yearly uWUE from \citet{Zhou_2015}, $g_1$ from \citet{Lin_2015} \citep[as presented in ][]{Franks_2017}, and observations of R, T, P, D$_s$, and wind speed (WS), the only unknown is $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$. With flux tower observations of ET, $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ will then be uniquely determined for each observation through Equation 6:

\begin{linenomath*}
  \begin{equation}
    \frac{\text{LAI}}{\text{LAI$_{ref}$}}  = - \frac{g_a \gamma c_s \sqrt{D_s} P }{ \left(\text{ ET } ( \Delta + \gamma) - \Delta R - g_a \rho_a c_p D_{s}\right) 1.6 \; R\; T\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{D_s}})}
    \label{lai}
  \end{equation}
\end{linenomath*}

%% see registers!!!!!! - help for emacs productivity
This $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ is some part a ``true'' deviation of  LAI from LAI$_{ref}$ in \citet{Zhou_2015}, and some part model and observational error, including assumptions about constant $g1$ and $uWUE$. By calculating a unique $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ for each observation we will propagate any model and observational uncertainty forward into our expression for $\frac{\partial \; ET}{\partial D}$.

\begin{table}
\caption{Definition of symbols and variables}
\centering
\begin{tabular}{l c c}
\hline
 Variable & Description & Units  \\
\hline
$e_s$  & saturation vapor pressure & Pa  \\ 
$T$  & temperature  & K \\
$\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ \\
$R$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   \\
  $g_a$  & atmospheric conductance & m s$^{-1}$  \\
  $\rho_a$  & air density & kg m$^{-3}$  \\
  $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ \\
  $D$  & VPD & Pa  \\
  $\gamma$  & psychrometric constant & Pa K$^{-1}$   \\
  $g_s$  & stomatal conductance & m s$^{-1}$  \\
  $g_{l-s}$  & leaf-scale stomatal conductance & mol m$^{-2}$ s$^{-1}$  \\
  $R*$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ \\
  $LAI$ & leaf area index & -\\
  $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ & ratio of LAI to LAI$_{ref}$ in \citet{Zhou_2015} & -\\
  $c_s$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air\\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

@@

By construction, the variability in the $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ term (Equation \ref{lai}) contains all model and observational uncertainties. We expect LAI to be cloase to LAI$_{ref}$, so $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ should be $O(1)$. We can have some confidence in our framework, including the assumption of constant uWUE, if calculated $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$s are generally $O(1)$. Figure \ref{lai_fig} presents the histogram of calculated $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$s with outliers (lowest and highest 5\% percent) and nonphysical values ($\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ $<$ 0.) removed. All remaining $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ values are $O(1)$ which provides confidence in model framework.

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig02.png}
\caption{Histogram of $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ values calculated for each site and time according to Equation \ref{lai}. The lowest and highest 5\% are removed as outliers, as well as any values below 0. The curve is normalized such that its area is 1. }
\label{lai_fig}
\end{figure}

An additional concern is that the $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ term may in fact be some function of $D$, in which case the dependence would need to be accounted for when taking the derivative. Figure \ref{lai_vpd_fig}\explain{Figs 2 and 3 can probably be combined - the histogram of $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ is shown in Fig 3}. plots the joint distribution of $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ and VPD, and shows that $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ is very weakly a function of VPD. Given this weak dependence, we argue that Equation \ref{d_et} is a valid approximation for ET response to $D$.

% below show histogram of site specific spearmannr? think would be useful 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $D$ and $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$. $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ has only a weak dependence on $D$. ***This plot could proabably benefit from a box plot of site specific correlations, because some sites do have stronger depednence than others. Note also Figs 3 and 2 can probably be combined because this figure shoes $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$'s histogram.***}
\label{lai_vpd_fig}
\end{figure}

Before calculating the sensitivity of ET to VPDn, it is useful to consider the functional form of Equation \ref{d_et}. There are three terms: a scaling term for the full expression we will call Term 1 ($\frac{g_a \; P}{T(\Delta + \gamma)}$), a relatively constant offset we will call Term 2 ($\frac{c_p}{R_{air}}$), and a variable term we will call Term 3 ($\frac{\text{LAI$_{ref}$}}{\text{LAI}}\frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{D}}{2 (g_1 + \sqrt{D})^2}\right)$). All variables are positive, so the relative magnitude between Term 2 and Term 3 will determine the sign of the derivative, while Term 1 will scale the expression larger or smaller.

@@

Term 2 minus Term 3 determines the sign of the sensitivty of ET to VPD and is thus crucial. $c_s$ variability is relatively less than $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ and $D$ variability, then variability within PFT will be solely determined by $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ and $D$ at the different fluxnet sites. Figure \ref{term3} shows how (Term 2 - Term 3) varies with $D$ and $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$, as a function of PFT. In Figure \ref{term3}a lower uWUE and $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ shift the distribution of (Term 2 - Term 3) towards negative values. Additionally, the smaller $g1$, the greater the relative $D$ dependence of of (Term 2 - Term 3). This is observed most strongly for the ENF PFT, which has the smallest $g1$ (74.31).

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for Term 2 - Term 3. Top: Term 2 - Term 3 as a function of VPD, with $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ held fixed at PFT averages. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$. Lines and stars denote the distribution of VPD and $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ next to each axis, following the same percentiles as above.}
\label{term3}
\end{figure}

Figure \ref{term3}b shows the location of the minima of ET, as a function of $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ and $D$. For any $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ or VPD less (more) than these curves, Term 2 - Term 3 will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$. However, we can see some general trends. For CSH, $\frac{\partial \; ET}{\partial \; D}$ should be negative for the vast majority of observed $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ and VPD. The fraciton of positive $\frac{\partial \; ET}{\partial \; D}$ appears to be more even for ENF, GRA, and DBF, and we might expect a greater frequency of positive $\frac{\partial \; ET}{\partial \; D}$ for CRO.

Table 3 confirms these expectations for PFT behavior of $\frac{\partial \; ET}{\partial \; D}$. For all PFTs except for CRO, average $\frac{\partial \; ET}{\partial \; D}$ is less than zero. However, $\frac{\partial \; ET}{\partial \; D}$ evaluated at the average of all variables (e.g. $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$, $T$, $c_s$, $D$) is only negative for CSH and GRA. And, DBF in addition to CRO experiences $\frac{\partial \; ET}{\partial \; D}$ < 0 less than half the time. These observations highlight the effect of the nonlinear function in Figure \ref{term3}: $\frac{\partial \; ET}{\partial \; D}$ has a much steeper slope when the function is negative, and is thus more likely to be large.

The units of $\frac{\partial \; ET}{\partial \; D}$ make it difficult to interpret if $D$ is even a meaningful contributor to ET's variability. To better understand $D$'s contribution, we normalize $\frac{\partial \; ET}{\partial \; D}$ with $D$'s standard deviation to define a (linearized) relative change in ET for variations in $D$ . $D$'s contribution to ET's variability ranges between 16 - 20 W m$^{-2}$ for all PFTs except for CSH, which is about 51 W m$^{-2}$. Another meaningful comparison is to $\frac{\partial \; ET}{\partial \; R} * std(R)$, as net radiationis generally the driver of ET (cite joe berry here). For all PFTs except for CSH $D$ contributes between 14.5 - 20.5 \% of $R$'s contribution to variability. For CSH the portion is much larger, about 44 \%. $D$'s variability is certiantly a non-negligable contributor to $ET$'s variability.

So far, idealized plots and statistics have illuminated the form of $\frac{\partial \; ET}{\partial \; D}$ and how it varies with PFT. Large mean $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ and uWUE shifts CRO and DBF towards positive $\frac{\partial \; ET}{\partial \; D}$. However, the strongly nonlinear function of $\frac{\partial \; ET}{\partial \; D}$ at $\frac{\partial \; ET}{\partial \; D} < 0$ pushes $\overline{\frac{\partial \; ET}{\partial \; D}}$ negative for DBF (it does not do this for CRO because of CRO's high $g1$). ENF's low $g1$ value increases the dependence of $\frac{\partial \; ET}{\partial \; D}$ on $D$, and makes the function more strongly nonlinear. This has the side effect of pushing $\overline{\frac{\partial \; ET}{\partial \; D}}$ negative further than other PFTs for a given fraction $\frac{\partial \; ET}{\partial \; D} < 0$ and magnitude $\frac{\partial \; ET}{\partial \; D}(\overline{T,\ldots,D})$. GRA shows the opposite behavior; a relatively high $g1$  makes the function more linear, decreasing the magnitude of $-\overline{\frac{\partial \; ET}{\partial \; D}}$ for a given  [large] fraction $\frac{\partial \; ET}{\partial \; D} < 0$ and negative $\frac{\partial \; ET}{\partial \; D}(\overline{T,\ldots,D})$ (although $g_a$ and Term 1 also probably have a role in this). Finally, low $uWUE$ of CSH pushes to toward by far the lowest values $\frac{\partial \; ET}{\partial \; D}$ (Figure \ref{term3}). Variability in $D$ accounts for the largest about of $ET$ variability for CSH. For the other PFTs, $D$ contributes less to $ET$ variability, but still represents about 15-20 \% of $R$'s contributions to ET variability.

@@

The full observations generally confirm expectations from Section \ref{results}. CRO has the most positive values of $\frac{\partial \; ET}{\partial \; D}$, $\frac{\partial \; ET}{\partial \; D}$ is almost always negative for CSH, and response depends more with the environmental conditions for the other PFTs (especially ENF). Through the columns of Figure \ref{real} we can see the impact of $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ and $g_a$ on the variability of $\frac{\partial \; ET}{\partial \; D}$. $g_a$'s scaling (included in columns 1 and 3) alters the magnitude considerably. $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ variability (included in columns 1 and 2) adds a lot of additional noise to the signal of $\frac{\partial \; ET}{\partial \; D}$, which is slightly undesirable given $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$'s role in representing model and observational uncertainty. However, the general story with the noise appears to match the cleaner signal when $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ is help constant and $D_{ETmin}$ is clearly visible \explain{I really need to make these plots better - way too much overlapping of points that hurts the story}. One exception is possibly with GRA, for which uncertainty represented in $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ is high and causes the full complexity plots (Columns 1 and 2) to not match well with $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ held fixed (Columns 3 and 4).

For ENF and GRA $D_{ETmin}$ does not appear to be only a function of $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ (most observable in Column 4). It turns out that the site to site variability in $\gamma$ causes $D_{ETmin}$ to vary, which is not discussed in the previous section. The impact is observable in both ENF and GRA, but especially for ENF which has a larger $\frac{\partial^2 \; ET}{\partial^2 \; D}$ than the other PFTs.

In general the full complexity plots of $\frac{\partial \; ET}{\partial \; D}$ match our expectations, even with the large sensitivity to $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ measures of uncertainty observed in Figure \ref{term3}. Our $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$-based method of uncertainty propagation blurs the idealized expectations the most for GRA, and also has a considerable effect for CRO. We therefor have the most confidence in our conclusion based on Equation \ref{d_et} for PFTS CSH, DBF, and ENF, as the full complexity plots with uncertainty included closely match the story when $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ is held fixed. **see somewhat preferred alternate figure \ref{real2} \explain{I think I like the alternate plot much  more as thinking in terms of T and RH is easier, and it makes the story easier to see at relatively low temperatures. However, I used the other plot because Fig 5 does not discuss things in terms of temperature, as this would make things more complicated (adding another dimension). I could just include both versions of figure 6 though (using figure 6a as a bridge to figure 6b)}.
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{./fig06.png}
\caption{Scatter plots of $\frac{\partial \; ET}{\partial \; D}$. Each row is a different PFT, and each column is a different quantity related to $\frac{\partial \; ET}{\partial \; D}$, as labeled: Column 1 - $\frac{\partial \; ET}{\partial \; D}$; Column 2 - $\frac{\partial \; ET}{\partial \; D}$ normalized by $g_a$; Column 3 - $\frac{\partial \; ET}{\partial \; D}$ with $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ held fixed at PFT average; and Column 4 - $\frac{\partial \; ET}{\partial \; D}$ normalized by $g_a$ and with $\frac{\text{LAI}}{\text{LAI$_{ref}$}}$ held fixed. For reference, lines corresponding to RH = 20\% and RH = 90 \% are drawn. Please note differences in the colorbar scale. ***see alternate (or additional) plot below.***}
\label{real}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{./fig06b.png}
\caption{****alternate Fig 06****  Scatter plots of $\frac{\partial \; ET}{\partial \; D}$. Each row is a different PFT, and each column is a different quantity related to $\frac{\partial \; ET}{\partial \; D}$, as labeled. If I end up using this, I could also draw on the curve of $D_{ETmin}$ with $\overline{$\frac{\text{LAI}}{\text{LAI$_{ref}$}}$}$. }
\label{real2}
\end{figure}
